\documentclass{article}
\author{Tyler Compton}
\title{Chapter 1}

\usepackage{float}

\begin{document}

\maketitle
\tableofcontents

\section{Introduction}
Computer Science is about how computing problems can be solved, Software
Engineering is about how humans can use computing to solve problems. It is the
study of the techniques and theory that support the development of high quality
software. This makes process very important.

Our goal as software engineers is to satisfy the client, who sponsors the
development of a product, and the user, who needs the product for a task.

A challenge of software engineering is solving the right problem. When talking
to clients, it is often unclear what problem they want solved. Oftentimes, a
client will describe how they would solve the problem, not what it is.

Software engineers are tasked to deliver a high quality solution that is
accomplished ethically. This must all be done using the allotted budget and
time.

\section{Quality Characteristics}
\begin{table}[H]
	\centering
	\begin{tabular}{ l l }
		Quality Characteristic & Description \\
		Correctness & How well the program adheres to the given requirements \\
		Reliability & The frequency and criticality of software failure \\
		Robustness & How well unexpected situations are handled \\
		Usability & How easy it is for the software to be used \\
		Maintainability & How quickly and easily changes can be applied to software \\
		Reusability & How easily the software components can be reused for future development \\
		Portability & How easily software can be used in multiple environments \\
		Efficiency & How much the software can do without wasting resources \\
	\end{tabular}
\end{table}

\section{Data Structures}
Generally, we talk about data structures as things that manage potentially
large collections. As an example, we can look at shipping containers on a dock.
There are potentially a lot of these containers, and they all need to be
managed quickly, efficiently, and none may be lost. Containers are generic
structures. They can hold all kinds of things, but the container itself is the
same. This allows handlers to not have to know what's inside. Each container
may have a unique identifier associated with it.

How should this problem be solved? How many containers will we have to deal
with at any given point? How will they be organized? Perhaps we need to sort
them. Does the order in which the containers is loaded or unloaded matter?

\end{document}
